---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# RUVpbps

<!-- badges: start -->
<!-- badges: end -->

RUVpbps create pseudosamples to implement the RUVIII method [Jacob et al., 2016] in pseudobulk analyses with insufficient technical replicates.

## Installation

You can install the development version of RUVpbps from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("esprietol/RUVpbps")

```
or

``` r
# install.packages("devtools")
devtools::install_github("esprietol/RUVpbps")

```



## Example

This is a basic example to generate pseudobulk pseudosamples (PBPS):

```{r example}
library(RUVpbps)
data(dummysce)

pbpsc <- gen_PBPS(dummysce, ctype='cg_cov', BioVar='cg_cov', NVar='Processing_Cohort', id_pb = 'sample_cell', id_sub = 'ind_cov', cell_id = 'cell_id', n = 1 )

```

## References

Jacob, L., Gagnon-Bartsch, J.-P., & Speed, T. P. (2016). Correcting gene expression data when neither the unwanted variation nor the factor of interest are observed. *Biostatistics*, 17(1), 16–28. https://doi.org/10.1093/biostatistics/kxv026
